Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-10-31 paper/manuscript.tex
removed ATM section, redistributed some content

While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, as discussed in Section \ref{subsec:numerical-implementations} it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs (Figure \ref{fig:atm_alchemical_intermediate})~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state (Figure \ref{fig:atm_alchemical_intermediate}). Current implementations use an alchemical intermediate state midway between the bound and unbound states with the symmetric hybrid potential energy function
\begin{equation}\label{eq:atm-intermediate-pot}
U_{1/2}(x) = \frac12 [ U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) + U(x_R, x_L, \zeta, \vec{r}_v)  ] 
\end{equation}
To include the alchemical intermediate, Eq.~(\ref{eq:atm-partfunc-ratio}) is multiplied and divided by the partition function of the system with the hybrid potential energy function in Eq.~(\ref{eq:atm-intermediate-pot}), each serving as the sampling state for each leg. The expression for the binding constant becomes a ratio of ensemble averages
\begin{equation}\label{eq:atm-average-abfe-intermediate}
  K_b = C^\circ V_{\rm site} \frac{
    \langle e^{-\beta \Delta U_1} \rangle_{1/2}
  }{
    \langle e^{-\beta \Delta U_2} \rangle_{1/2}
  }
\end{equation}
where $\Delta U_1$ and $\Delta U_2$ are the perturbation energies of the bound and unbound states relative to the alchemical intermediate. For example,
\begin{equation}
\Delta U_1(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U_{1/2}(x)
\end{equation}
Setting,
\begin{equation}
\Delta G_1 = -k_B T \ln \langle e^{-\beta \Delta U_1} \rangle_{1/2}
\end{equation}
and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling. This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer. However, alchemical transfer simulates both legs with a single simulation system while double-decoupling accomplishes them in separate simulations.

@@

In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occurs at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{azimi2022relative} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other). %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{ azimi2022relative} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

Note that $\Delta G^\circ_{b,{\rm ideal}}$ is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing similar binding site definitions. The latter is the reason that the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs). However, while it does not appear explicitly in the definition, $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$. Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}. Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions. Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined. The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.

@@

There are two main classes of approaches to address artefacts introduced by changes in net charge during alchemical free energy calculations: (1) avoiding changing the net charge altogether and (2) post-processing approaches to correct for artefacts using continuum electrostatics.
Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}. In absolute free energy calculations, it is often difficult to avoid a change in the net charge. One way to avoid charge changes is to ensure that, as a charged ligand is removed, a charged counterion of opposite sign is also removed, or one of the same sign is inserted. This is sometimes referred to as a “co-alchemical ion” approach to deal with the required charge change. This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
There are special flavours of the co-alchemical ion approach such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup. Here, the two branches of a thermodynamic cycle as used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation. Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.

@@

@@

Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}. For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerated,sidler2017efficient, christ2007enveloping}. When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation. However, it can often fail whenever the simulation of this intermediate state ends up trapped in configurations characteristic of only one end state. Thus, successful use of EDS can require system-dependent tuning, making it difficult to implement in an automated and reliable way. However, when successful, it can be very efficient.

@@

\subsubsection{What $\lambda$-sampling schemes are available?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{lee2023aces}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

The potential energy differences required for EXP, BAR, MBAR, and WHAM can be calculated either during the simulation or in post-processing. It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by evaluating the energy of the configuration twice, and to reduce the amount of stored information. 
Although potential energy derivatives must usually be calculated in code there is one condition under which they can be easily computed in post-simulation analysis. 
If the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_0(\vec{q}) - u_1(\vec{q})$, which is the difference between the initial and final states, which are already calculated by the simulation and can be recorded easily without additional computational expense. 
However, because of the problems with linear paths already discussed in this paper, this simplification is rarely that useful.

@@

\subsection{Alchemical Transfer Method for Absolute and Relative Binding Free Energy Calculations}
\label{sec:atm}
\egnote{This section will be removed and some of the content dispersed among the relevant sections of the paper}

\subsubsection{Introduction}

The Alchemical Transfer Method (ATM)~\cite{wu2021alchemical,azimi2022relative} is an alchemical framework for estimating binding free energies that targets generality with a rigorous and straightforward implementation, with as few additions as possible to standard non-alchemical models of molecular systems. Additionally, the method targets complex ligand transformations in drug design, involving scaffold-hopping and charge-changing transformations that are not well supported by traditional algorithms. ATM's overarching philosophy is to deploy robust and rigorous free energy estimation algorithms with the simplest possible procedures for developers and end users. One critical design principle is the seamless support for advanced molecular potential energy functions, such as quantum-mechanical, polarizable, and neural network potentials, in free energy-based drug discovery applications.

To reach its objectives, ATM represents the alchemical potential energy function through a coordinate transformation~\cite{azimi2022application,azimi2022relative}. Unlike traditional alchemical approaches that rely on parameter interpolation of interaction potentials, ATM employs an energy interpolation alchemical energy function~\cite{konig2021efficient} where the target potential is expressed as the reference potential after the ligand is translated from the solvent to the binding site (Figure~\ref{fig:atm-illustration}). The binding free energy calculation is performed directly, without solution and receptor legs, as well as electrostatic and non-electrostatic steps. The simulation system consists of a single solvent box prepared with conventional setup tools and without pairwise soft-core potentials.  The leveraging of the natural capabilities of MD engines to apply coordinate transformations is a distinguishing feature of ATM. Alchemical transformations for binding implemented through coordinate transformations are computationally straightforward and applicable with any potential energy function, requiring only the evaluation of energy and derivatives before and after the coordinate transformation. Conversely, because it recalculates the potential energy and its gradients twice at each time-step, ATM's strategy suffers from the downside that it is as much as twice as slow per time-step as conventional parameter interpolation alchemical approaches.

%mentions parameter vs. energy interpolation 
%@article{konig2021efficient,
%  title={Efficient Alchemical Intermediate States in Free Energy Calculations Using $\lambda$-Enveloping Distribution Sampling},
%  author={K\"{o}nig, Gerhard and Ries, Benjamin and H\"{u}nenberger, Philippe H and Riniker, Sereina},
%  journal={Journal of Chemical Theory and Computation},
%  volume={17},
%  number={9},
%  pages={5805--5815},
%  year={2021},
%  publisher={ACS Publications}
%}

\subsubsection{Methodology}

Using the notation in Eq.~(\ref{eq:reduced-potential}), if $u(\vec{q})$ is the physical ($\lambda$-independent) reduced potential of the system when the ligand is at a location in the solvent away from the receptor and $\vec{h}$ is a displacement vector that brings the ligand atoms from the position in the solvent to the receptor binding site (Figure~\ref{fig:atm-illustration}), the ATM \emph{reduced binding energy function}
\begin{equation}\label{eq:ATM-binding-energy}
b(\vec{q}) = u(\vec{q}+\vec{h}) - u(\vec{q})
\end{equation}
is the change in the system's potential energy when the ligand is rigidly moved from the solvent to the receptor binding site. The ATM alchemical reduced potential is then written as
\begin{equation}\label{eq:ATM-alchemical-potential}
    u(\vec{q}; \vec{\lambda}) = u(\vec{q}) + w[b(\vec{q});\vec{\lambda}]
\end{equation}
where $w(b,\vec{\lambda})$ is the \emph{perturbation energy function}. The perturbation energy function can be any function as long as it is zero at the $\lambda$-state corresponding to the unbound state and it is equal to the binding energy $b(\vec{q})$ at the bound state:
\begin{eqnarray}
    w[b(\vec{q});\vec{\lambda}_{\rm U}] = 0           & \quad & \vec{\lambda}_{\rm U}: {\rm Unbound\ State} \\
    w[b(\vec{q});\vec{\lambda}_{\rm B}] = b(\vec{q})  & \quad & \vec{\lambda}_{\rm B}: {\rm Bound\ State}
\end{eqnarray}

Current ATM implementations (see below) employ the softplus perturbation energy function\cite{pal2019perturbation}
\begin{equation}\label{eq:ATM-softplus}
w(b;\vec{\lambda}) = w[b_{\rm sc}(b);\vec{\lambda}] =  \frac{\lambda_2 - \lambda_1}{\alpha} \ln\left\{ 1 + e^{-\alpha [b_{\rm sc} - b_0]} \right\} + \lambda_2 b_{\rm sc}
\end{equation}
where the parameters ($\lambda_1>=0$, $\lambda_2>=0$, $\alpha>0$, and $b_0$) are functions of the $\lambda$-state, $\vec{\lambda}$. The softplus non-linear perturbation function satisfies the correct limits when $\lambda_1 = \lambda_2 = 0$ at the unbound state and $\lambda_1 = \lambda_2 = 1$ at the bound state. The quantity $b_{\rm sc}$ in Eq.~(\ref{eq:ATM-softplus}) is the soft-core binding energy function
\begin{equation}
  b_{\rm sc}(b)=
\begin{cases}
b & b \le b_c \\
(b_{\rm max} - b_c ) f_{\rm sc}\left[\frac{b-b_c}{b_{\rm max}-b_c}\right] + b_c & b > b_c
\end{cases}
\label{eq:ATM-soft-core-general}
\end{equation}
with
\begin{equation}
f_{\rm sc}(y) = \frac{z(y)^{a}-1}{z(y)^{a}+1} \label{eq:rat-sc} \, ,
\end{equation}
and
\begin{equation}
    z(y)=1+2 y/a + 2 (y/a)^2
    \label{eq:end-of-softplus}
\end{equation}
The soft-core perturbation energy function is designed to avoid singularities near the initial state of the alchemical transformation.\cite{pal2019perturbation,khuttan2021alchemical} The parameters $b_{\rm max}$, $b_c$, and $a$ are set to smoothly cap the perturbation energy to a maximum positive value without affecting it away from the singularity. 
The softplus function has the nice property that is linear for $b<<b_0$ and $b>>b_0$ but with different slopes,  $\lambda_1$ and $\lambda_2$, in these two regimes. The bilinear property of the softplus function allows for greater flexibility in creating smoother and more connected alchemical paths that lead to faster convergence~\cite{pal2019perturbation, khuttan2021alchemical}.

The values of the binding energies sampled at regular intervals at each $\lambda$-state during the ATM alchemical simulation are the inputs for binding free energy estimation (see Section \ref{sec:data_analysis}) through multi-state thermodynamic reweighting algorithms such as MBAR and UWHAM~\cite{shirts2008statistically,Tan2012}. One requirement for these algorithms is the evaluation of the perturbation energy $w(b,\vec{\lambda})$ for each sample $b$ at all $\lambda$-states.  The analytical relationships Eq.~(\ref{eq:ATM-softplus})--(\ref{eq:end-of-softplus}) between the binding energy and the perturbation energy make this step straightforward with ATM. In contrast, conventional free energy perturbation methods that employ soft-core pair potentials lack analytical perturbation energy functions and require a more complex process involving calling the molecular dynamics engine to rescore simulation trajectories. The more straightforward free energy estimation process afforded by ATM saves computing and storage costs, and, since it is based solely on binding energies, allows for changing the perturbation energy function $w(b;\vec{\lambda})$ without running new simulations. The simplification of the free energy estimation process fits the general ATM philosophy of streamlining and generalizing alchemical binding free energy calculations as much as possible.

>>>>

While Eq.~(\ref{eq:atm-average-abfe}) is formally rigorous, it is numerically necessary to divide the alchemical transfer ABFE calculation in explicit solvent into two legs~\cite{wu2021alchemical,azimi2022relative} that connect the unbound and bound states by a suitable alchemical intermediate state. Current implementations use an alchemical intermediate state midway between the bound and unbound states with the symmetric hybrid potential energy function
\begin{equation}\label{eq:atm-intermediate-pot}
U_{1/2}(x) = \frac12 [ U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) + U(x_R, x_L, \zeta, \vec{r}_v)  ] 
\end{equation}
To include the alchemical intermediate, Eq.~(\ref{eq:atm-partfunc-ratio}) is multiplied and divided by the partition function of the system with the hybrid potential energy function in Eq.~(\ref{eq:atm-intermediate-pot}), each serving as the sampling state for each leg. The expression for the binding constant becomes a ratio of ensemble averages
\begin{equation}\label{eq:atm-average-abfe-intermediate}
  K_b = C^\circ V_{\rm site} \frac{
    \langle e^{-\beta \Delta U_1} \rangle_{1/2}
  }{
    \langle e^{-\beta \Delta U_2} \rangle_{1/2}
  }
\end{equation}
where $\Delta U_1$ and $\Delta U_2$ are the perturbation energies of the bound and unbound states relative to the alchemical intermediate. For example,
\begin{equation}
\Delta U_1(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U_{1/2}(x)
\end{equation}
Setting,
\begin{equation}
\Delta G_1 = -k_B T \ln \langle e^{-\beta \Delta U_1} \rangle_{1/2}
\end{equation}
and similarly for the second leg, we recover Eq.~(\ref{eq:ddm-dg1-dg2}) obtained for double-decoupling. This shows that, while the thermodynamic path and the meaning of the $\Delta G_1$ and $\Delta G_2$ free energy components differ, the double-decoupling and alchemical transfer formulations both connect the unbound and bound states through an auxiliary intermediate state: the ligand decoupled state for double-decoupling, and the symmetric alchemical intermediate for alchemical transfer.~\cite{azimi2025potential} However, alchemical transfer simulates both legs with a single simulation system, while double-decoupling accomplishes them in separate simulations.

@@

As discussed below, because of poor ensemble overlap, in most cases of interest, alchemical free energy calculations cannot be carried out directly by ensemble averages collected only at the reference or target states. Instead, in a process called stratification,~\cite{chipot2007free} we take advantage of the state function property of the free energy and break up the transformation into a series of manageable steps, distributed along an alchemical pathway, whose combination is equivalent to the desired transformation (see Section \ref{sec:important_path}). Intermediate alchemical states are typically defined in terms of one or more alchemical progress parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential energies of the initial and final states are recovered at two particular values $\vec{\lambda}_0$ and $\vec{\lambda}_1$.

In the case of a relative binding free energy calculation, for example, this might be achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$. A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occurs at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{azimi2022relative} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other). %In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand.
In ATM, restraints are used to keep the bound and free ligands at some distance apart and in the same orientation, so that during the translation move the free ligand is placed into a position and orientation as similar as possible to that of the bound ligand. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

Because they duplicate the common region, dual coordinate formulations are not generally applicable to small modifications of large molecules, such as single-point mutations of proteins. An exception, and an example of the variety of alchemical approaches, is the Alchemical Transfer with Coordinate Swapping method~\cite{gallicchio2025mutants}, which is a dual coordinate approach with similar convergence characteristics of single- and dual-topology formulations. It works by exchanging the coordinates of the corresponding atoms of the common substructure of the free and bound ligands and displacing the atoms of the variable substituents. The net effect is the switching of the positions of the two ligands (Figure \ref{fig:atm-illustration}\textbf{B}) by the transfer of only their variable regions.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{azimi2022relative} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

Note that the ideal binding free energy term $\Delta G^\circ_{b,{\rm ideal}}$ in Eq.~(\ref{eq:DG0-def-Vsite}) is independent of the chemical nature of the binding partners, and it is the same for all complexes sharing binding site definitions. The latter is why the ideal term cancels out when taking the difference of absolute binding free energies when estimating relative binding free energies (RBFEs). However, while it does not appear explicitly in its definition in Eq.~(\ref{eq:DG0-def-Vsite}), the excess component of the binding free energy $\Delta G_{b,{\rm exc.}}$ depends implicitly on the definition of the binding site through the ensemble average in the uncoupled ensemble, which includes only ``bound'' configurations according to the definition of the indicator function $I(\zeta)$. Varying the binding site definition affects both the ideal and excess binding free energy terms such that their sum (the standard binding free energy) is often only weakly dependent on the binding site definition~\cite{gilson1997statisticalthermodynamic,gallicchio2011recent}. Conversely, great care should be exercised when comparing calculated excess binding free energy values corresponding to different binding site definitions. Similarly, the outcome of a binding free energy calculation conducted without a binding site definition, i.e.\ without binding site restraints, is, in principle, undefined. The latter is also true for relative binding free energies because while the ideal term cancels out, the difference in excess binding free energies depends on the binding pose definition.

@@

There are two main classes of approaches to address artefacts introduced by changes in net charge during alchemical free energy calculations: (1) avoiding changing the net charge altogether and (2) post-processing approaches to correct for artefacts using continuum electrostatics.
Many relative free energy planning tools have been set up to avoid changing the net charge of the systems considered, including LOMAP~\cite{liu2013lead} and Schr\"{o}dinger's FEP+~\cite{wang2015accurate}. In absolute free energy calculations, it is often difficult to avoid a change in the net charge. One way to avoid charge changes is to ensure that, as a charged ligand is removed, a charged counterion of opposite sign is also removed, or one of the same sign is inserted. This is sometimes referred to as a “co-alchemical ion” approach to deal with the required charge change. This approach is easily supported in most free energy calculating programs, as it merely involves an additional simultaneous modification.
There are special flavours of the co-alchemical ion approach, such as proposed by Gapsys et al.~\cite{gapsys2015calculation}, which uses a double-system/single-box setup. Here, the two branches of a thermodynamic cycle used in relative free energy calculations are placed in one simulation box, effectively circumventing a change in the overall net charge during a perturbation. 
Note that the co-alchemical approaches only avoid finite-size effects with lattice-sum methods such as PME.
The Alchemical Transfer approach,\cite{wu2021alchemical,azimi2022relative} discussed in Section \ref{subsec:formulations-abfe-rbfe}, is naturally free of charge-changing artefacts because it encodes alchemical binding processes by coordinate transformations that conserve the system's net charge.

@@

The pathway implemented for the Alchemical Transfer Method (ATM)~\cite{pal2019perturbation,khuttan2021alchemical,wu2021alchemical,azimi2022relative} is an example of an alternative alchemical formulation for binding that does not require soft-core pair potentials and the splitting of electrostatic and non-electrostatic contributions. ATM employs an optimized softplus alchemical potential energy function~\cite{pal2019perturbation} and a soft-core binding energy function that smooths the variations of the perturbation energy $u(\vec{q}) = U_1(\vec{q}) - U_0(\vec{q})$ rather than each pair interaction.\cite{khuttan2021alchemical} The advantage of this approach is that it does not require modifications of the energy routines of the MD engine to implement soft-core pair potentials. Moreover, it has been shown to be applicable without electrostatic/LJ splitting to the transfer of molecules, equivalent to the disappearance and introduction of atoms in conventional single- and dual-topology methods.\cite{khuttan2021alchemical}

@@

Additional approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}. For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerated,sidler2017efficient, christ2007enveloping}. When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation. However, it can often fail whenever the simulation of this intermediate state ends up trapped in configurations characteristic of only one end state. Thus, successful use of EDS can require system-dependent tuning, making it difficult to implement in an automated and reliable way. However, when successful, it can be very efficient.

@@

\subsubsection{What $\lambda$-sampling schemes are available?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for some AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization},
AMBER~\cite{Lee_JChemInfModel_2020_v60_p5595}, FEP+~\cite{wang2015accurate}, NAMD~\cite{jiang2019computing}, and OpenMM~\cite{rufa2020chemical,azimi2022relative}.
A specific variant of Hamiltonian replica exchange is FEP/REST2~\cite{rest2, feprest} that involves ``heating'' a small portion of the protein-ligand system to enhance the sampling within the binding pocket. It has been promoted as a great way to enhance sampling and is employed by default in Schr\"{o}dinger's FEP+ package. However, it should be noted that several studies have questioned its reliability and reported that it degrades the quality of free energy predictions as compared to those based on standard MD simulations without any enhanced sampling~\cite{wan2020fep+, bhati2022}. Therefore, a blind application of this sampling scheme is not recommended and it must be used with caution if at all. A variant of this method that overcomes some of the reported issues with FEP/REST2 is called alchemical enhanced sampling (ACES)~\cite{lee2023aces}. ACES creates localized targeted "enhanced sampling states" that are connected to the real state end-points through HREMD.  The method leverages the dual topology framework to produce a concerted “counter-diffusion” of alchemical states that leads to minimal rearrangement of the environment along the alchemical path~\cite{York_ACSPhysChemAu_2023_v3_p478}. The computational overhead of the method is negligible relative to more conventional alchemical free energy simulations with HREMD.

@@

The potential energy differences required for EXP, BAR, MBAR, and WHAM can be calculated either during the simulation or in post-processing. It is recommended to calculate the potential differences in code when possible to avoid extra overhead and possible errors produced by evaluating the energy of the configuration twice, and to reduce the amount of stored information. 
Although potential energy derivatives must usually be calculated in code, they can sometimes be easily computed in post-simulation analysis. 
For example, if the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_1(\vec{q}) - u_0(\vec{q}) = u(\vec{q})$, which is the perturbation energy $u$--the potential energy difference between the initial and final states. These values are already calculated by the simulation and can be recorded easily without additional computational expense. As discussed earlier, linear paths are inefficient. However, methods such as EDS and ATM, based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values. The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).

For example, if the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_1(\vec{q}) - u_0(\vec{q}) = u(\vec{q})$, which is the perturbation energy $u$--the potential energy difference between the initial and final states. These values are already calculated by the simulation and can be recorded easily without additional computational expense. As discussed earlier, linear paths are inefficient. However, methods such as EDS and ATM based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values. The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).

@@
